---
title: "RMarkdown Exercise"
author: "UCI DataScience Initiative"
date: "March 04, 2016"
output: html_document
---

```{r warning=FALSE, message=FALSE, echo=FALSE, warning=FALSE}
library(knitr)
```


# Data Exploration

## Data Exploration using mtcars dataset:
1) mtcars dataset is available in R base package. Using str, summarize thi object:

```{r }
str(mtcars) # or ?mtcars
```

2) What are the dimesions? Explain in words whre dimensions are inline R codes!

```{r }
# install.packages(repos = "http://cran.stat.ucla.edu", "car")
dim(mtcars)
colnames(mtcars) # names(mtcars)
```

3) Take a subset of the data with these attributes:
  * "mpg", "cyl", "hp", "wt", "am", "gear"

```{r}
# we can take a subset of the data:
mtcars_sub <- mtcars[,c("mpg", "cyl", "hp", "wt", "am", "gear")]
```

4) Provide some useful descriptive statistics from the data:

```{r}
# About the data:
summary(mtcars_sub$mpg)
summary(mtcars_sub)

# for discrete attributes:
table(mtcars_sub$gear)
NumCyl <- table(mtcars_sub$cyl)
NumCyl

# two-way freq tables:
table(mtcars_sub$am, mtcars_sub$gear) # am refers to the manual transmission

freqTable <- table(mtcars_sub$am, mtcars_sub$gear)
rownames(freqTable) <- c("Automatic", "Manual")
colnames(freqTable) <- paste(3:5, "gears", sep = "-")
freqTable
kable(freqTable, caption = "Freq Table of Transmission vs. Number of Forward gears")
```

5) Create a histogram of horse-power:

```{r warning=FALSE, message=FALSE, error=FALSE}
# library(ggplot2) # http://www.cookbook-r.com/Graphs/
# ggplot(mtcars_sub, aes(as.factor(cyl))) + geom_histogram(binwidth = 1)
hist(mtcars_sub$hp, main = "hist of number of Horse Power")
```
