---
title: "ggplot2 Introduction"
author: "Lingge Li"
date: "12/2/2016"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## ggplot2 looks awesome

- The beautiful style of ggplot2 graphics perhaps is most people's first impression 

```{r echo=FALSE, eval=TRUE, warning=FALSE, fig.width=4, fig.height=3}
library(ggplot2)
ggplot(data=iris, aes(x=Species, y=Petal.Width)) +
  geom_violin(aes(fill=Species)) +
  theme(legend.position='none') +
  labs(title='Iris petal width by species')
```

## Base plot

```{r iris, echo=FALSE, eval=TRUE, fig.width=5, fig.height=4}
boxplot(iris$Petal.Width~iris$Species, 
        main='Iris petal width by species')
``` 

## However...

- On second glance, the code seems slightly complicated

```{r echo=TRUE, eval=FALSE}
ggplot(data=iris, aes(x=Species, y=Petal.Width)) +
  geom_violin(aes(fill=Species)) +
  theme(legend.position='none') +
  labs(title='Iris petal width by species')
```

## Layered grammar of graphics

- ggplot2 follows a specific grammar of graphics

\includegraphics[width=300pt]{layer.png}

*Example taken from Hadley Wickham's book*

http://vita.had.co.nz/papers/layered-grammar.pdf

## How to make a plot

- Data
- Geometric objects (geom)
- Aesthetic mapping (aes)
- Statistical transformation (stat)
- Scales and coordinate system

## Geom

- Wide range of geometric objects from basic points (geom_point) to statistical plots (geom_violin)

- Add geometric objects with +

- Multiple geometric objects on the same plot

## Data + mapping

- Must supply data and aes()

- Dataframe and xy positions always needed

- Each geometric object can have its own

```{r echo=TRUE, eval=FALSE}
geom_point(data, aes(x, y))
```

## Scatterplot

```{r echo=TRUE, eval=TRUE, fig.width=3, fig.height=2}
ggplot() +
  geom_point(data=iris, 
             aes(x=Petal.Length, y=Petal.Width))
```

```{r echo=TRUE, eval=FALSE, fig.width=3, fig.height=2}
ggplot(data=iris, aes(x=Petal.Length, y=Petal.Width)) +
  geom_point()
```

## Aesthetics
- Other aesthetic mapping arguments include colour, fill, shape, size...

- How would you change the shape

```{r echo=TRUE, eval=TRUE, fig.width=4, fig.height=2}
ggplot(data=iris, aes(x=Petal.Length, y=Petal.Width)) +
  geom_point(aes(colour=Species))
```

## Mapping vs setting

- What's the difference here

```{r echo=TRUE, eval=TRUE, fig.width=3, fig.height=2}
ggplot(data=iris, aes(x=Petal.Length, y=Petal.Width)) +
  geom_point(colour='red')
```

## Boxplot

- How would you use geom_boxplot to create this

```{r echo=FALSE, eval=TRUE, fig.width=4, fig.height=2}
ggplot(data=iris, aes(x=Species, y=Petal.Width)) +
  geom_boxplot(aes(fill=Species))
```

## Layers

- Violin plots over boxplots

```{r echo=TRUE, eval=TRUE, fig.width=4, fig.height=2}
p <- ggplot(data=iris, aes(x=Species, y=Petal.Width)) +
  geom_boxplot(alpha=0.5)

p + geom_violin(aes(fill=Species), alpha=0.5)
```

## Motivating example

\includegraphics[width=200pt]{moma.png}

http://fivethirtyeight.com/features/a-nerds-guide-to-the-2229-paintings-at-moma/

## MoMA data

- Original dataset contains the entire MoMA collection

https://github.com/MuseumofModernArt/collection

- 2267 paintings by 989 artists

- Many interesting variables

## Height vs width

- Pre-process the data by removing outliers and missing values

```{r echo=TRUE, eval=TRUE}
Paintings <- read.csv("Paintings.csv")
Paintings <- Paintings[Paintings$Height..cm.<500,]
Paintings <- Paintings[Paintings$Width..cm.<1000,]
Paintings <- Paintings[!is.na(Paintings$Height..cm.),]
Paintings <- Paintings[!is.na(Paintings$Width..cm.),]
```

```{r echo=FALSE, eval=TRUE, fig.width=3, fig.height=2}
ggplot(data=Paintings, aes(x=Width..cm., y=Height..cm.)) + geom_point(alpha=0.4)
```

## Color by orientation

- How can we create a new variable in the data frame

```{r echo=FALSE, eval=TRUE, fig.width=4, fig.height=2}
Paintings$landscape <- Paintings$Width..cm. > Paintings$Height..cm.

ggplot(data=Paintings, aes(x=Width..cm., y=Height..cm.)) + 
  geom_point(aes(colour=landscape), alpha=0.4)
```

## Regression line

- Plot the regression line of height~width with geom_line() (Hint: lm(Y~X)$fitted.values)

```{r echo=FALSE, eval=TRUE, fig.width=4, fig.height=2}
fit <- lm(Paintings$Height..cm.~Paintings$Width..cm.)
Paintings$fitted <- fit$fitted.values

p <- ggplot(data=Paintings, aes(x=Width..cm., y=Height..cm.)) + 
  geom_point(aes(colour=landscape), alpha=0.4)
p + geom_line(aes(x=Width..cm., y=fitted))
```

## Separate lines

- How can we draw two separate regression lines (Hint: use different subsets of the data)

```{r echo=FALSE, eval=TRUE, fig.width=4, fig.height=2}
Landscape <- Paintings[Paintings$landscape,]
fit <- lm(Landscape$Height..cm.~Landscape$Width..cm.)
Landscape$fitted <- fit$fitted.values

Portrait <- Paintings[!Paintings$landscape,]
fit <- lm(Portrait$Height..cm.~Portrait$Width..cm.)
Portrait$fitted <- fit$fitted.values

p <- ggplot(data=Paintings, aes(x=Width..cm., y=Height..cm.)) + 
  geom_point(aes(colour=landscape), alpha=0.4)

p <- p + geom_line(data=Landscape, aes(x=Width..cm., y=fitted))

p + geom_line(data=Portrait, aes(x=Width..cm., y=fitted))
```

## Smoother

- There is actually a much easier way with geom_smooth()

```{r echo=TRUE, eval=TRUE, fig.width=4, fig.height=2}
p <- ggplot(data=Paintings, 
            aes(x=Width..cm., y=Height..cm.)) + 
  geom_point(aes(colour=landscape), alpha=0.4)

p + geom_smooth(aes(colour=landscape), 
                method='lm', se=FALSE)
```

## Scales

- Maybe the unit should be inch instead of centimeter (Hint: http://docs.ggplot2.org/0.9.3/scale_continuous.html)

```{r echo=FALSE, eval=TRUE, fig.width=4, fig.height=2}
ggplot(data=Paintings, aes(x=0.393701*Width..cm., y=0.393701*Height..cm.)) + 
  geom_point(aes(colour=landscape), alpha=0.4) +
  geom_smooth(aes(colour=landscape), method='lm', se=FALSE) +
  scale_x_continuous(name='Width (inches)') +
  scale_y_continuous(name='Height (inches)')
```

## Details

- We can change many other details (http://docs.ggplot2.org/0.9.2.1/theme.html)

```{r echo=TRUE, eval=FALSE}
ggplot(data=Paintings, 
       aes(x=0.393701*Width..cm., y=0.393701*Height..cm.)) + 
  geom_point(aes(colour=landscape), alpha=0.4) +
  geom_smooth(aes(colour=landscape), method='lm', se=FALSE) +
  scale_x_continuous(name='Width (inches)') +
  scale_y_continuous(name='Height (inches)') +
  theme_bw() + theme(legend.position='none') +
  annotate('point', x=0.393701*240, y=0.393701*200, 
           colour='blue', alpha=0.6) + 
  annotate('text', x=0.393701*400, y=0.393701*200, 
           label='Some famous art') +
  ggtitle('MoMA Painting Dimensions')
```

## End result

```{r echo=FALSE, eval=TRUE, fig.width=5, fig.height=4}
ggplot(data=Paintings, 
       aes(x=0.393701*Width..cm., y=0.393701*Height..cm.)) + 
  geom_point(aes(colour=landscape), alpha=0.4) +
  geom_smooth(aes(colour=landscape), method='lm', se=FALSE) +
  scale_x_continuous(name='Width (inches)') +
  scale_y_continuous(name='Height (inches)') +
  theme_bw() + theme(legend.position='none') +
  annotate('point', x=0.393701*240, y=0.393701*200, 
           colour='blue', alpha=0.6) + 
  annotate('text', x=0.393701*400, y=0.393701*200, 
           label='Some famous art') +
  ggtitle('MoMA Painting Dimensions')
```

## Facet

- We can make similar plots for different regions with facet

```{r echo=TRUE, eval=TRUE, fig.width=4, fig.height=2}
Paintings$American <- ifelse(Paintings$Nationality=='American', 
                             'American', 'Non-American')
ggplot(data=Paintings, aes(x=Width..cm., y=Height..cm.)) + 
  geom_point(alpha=0.4) +
  geom_smooth(method='lm', se=FALSE) +
  facet_grid(.~American)
```

## Multiple plots

- Multiplot function 

http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/  

- gridExtra package 

https://cran.r-project.org/web/packages/gridExtra/vignettes/arrangeGrob.html

## Last comment

- I hope you have gained a better understanding of ggplot2. There are plenty of tutorials and other resources online.

https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf

http://www.cookbook-r.com/Graphs/

http://stackoverflow.com/questions/tagged/ggplot2