---
title: "Interactive Plots with R Shiny"
author: "Lingge Li"
date: "March 4, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## An inspiring example

http://shiny.rstudio.com/gallery/superzip-example.html

What we are building today is not quite as fancy but shares the same kind of structure.

http://54.201.252.124:3838/linggeli7/Interactive/

## A static plot

The following code will produce a static scatter plot. In this case, we want to overlay the points on top of a map with the maps library. The maps library uses ggplot to display geographical regions as polygons. There's also the [ggmap](https://journal.r-project.org/archive/2013-1/kahle-wickham.pdf) library that uses google maps.

```{r eval=FALSE}
library(shiny)
library(ggplot2)
library(maps)

shinyServer(function(input, output) {
  long <- runif(20, min=-122.7, max=-115)
  lat <- runif(20, min=32.5, max=42)
  mydata <- data.frame(long, lat, replicate(50, rnorm(20)))
  
  all_states <- map_data("state")
  california <- all_states[which(all_states$region == "california"),]
  
  output$plot1 <- renderPlot({
    p <- ggplot()
    p <- p + geom_polygon( data=california, aes(x=long, y=lat, group = group),colour="white", fill="orange" ,alpha = 0.2)
    p <- p + geom_point( data=mydata, aes(x=long, y=lat), size=3, color="red")  
    p
  })
  
})

library(shiny)
shinyUI(fluidPage(
  plotOutput("plot1")
))
```

## Mouse events
In general, interaction happens through mouse events in three steps. First, we need to monitor mouse events. The mouse events are: click, double-click, hover, and brush. Then we will retrieve the values of the mouse events. Lastly, we will change the output accordingly. With R Shiny, it is very easy to keep track of the mouse events by including a few arguments in the UI. Their values are passed to the server in the input.
```{r eval=FALSE}
plotOutput("plot1", click = "plot_click",
                  dblclick = "plot_dblclick",
                  hover = "plot_hover",
                  brush = brushOpts(
                    id = "plot_brush",
                    resetOnNew = TRUE)),
fluidRow(
    column(width = 3,
           verbatimTextOutput("click_info")
    )
  )

output$click_info <- renderPrint({
    cat("input$plot_click:\n")
    str(input$plot_click)
  })
```
**Exercise 1** Can you print out hover, double-click and brush as well? What do you notice in the output?

## Zoom with double-click and brush
Now we will implement zoom with double-click and brush using [the reactive programming model](http://shiny.rstudio.com/articles/reactivity-overview.html).
```{r eval=FALSE}
ranges <- reactiveValues(x = NULL, y = NULL)

observeEvent(input$plot_dblclick, {
    brush <- input$plot_brush
    if (!is.null(brush)) {
      ranges$x <- c(brush$xmin, brush$xmax)
      ranges$y <- c(brush$ymin, brush$ymax)
      
    } else {
      ranges$x <- NULL
      ranges$y <- NULL
    }
  })
```
**Exercise 2** How can you alter the plot to zoom? Hint: [coord_cartesian()](http://docs.ggplot2.org/current/coord_cartesian.html).

## Click and show
We can find the closest point in the dataset to where has been clicked with [nearPoints](http://shiny.rstudio.com/reference/shiny/latest/nearPoints.html).
```{r eval=FALSE}
values <- reactiveValues(point = NULL)

observeEvent(input$plot_click, {
    current <- nearPoints(mydata, input$plot_click, "long", "lat", threshold = 10, maxpoints = 1)
    if (nrow(current) != 0) {
      values$point <- current
    } else {
      values$point <- NULL
    }
  })

```
Now that we have the entire row of data associated with the point. Here I use the data to make a line plot. You can make a histogram, a table or whatever you like.
```{r eval=FALSE}
output$plot2 <- renderPlot({
    if (is.null(values$point)) {
    } else {
      series <- data.frame(seq(50), t(values$point[3:52]))
      rownames(series) <- NULL
      colnames(series) <- c("time", "obs")
      ggplot(series, aes(x=time, y=obs)) + geom_line() 
    }
  })

plotOutput("plot2")
```
**Exercise 3** Do something creative with the data.

## What's next

Now you should have a good understanding of how interactive plots work. There are [several other things](http://shiny.rstudio.com/gallery/plot-interaction-advanced.html) you can do. You can also make plots look cooler with [HTML widgets](http://www.htmlwidgets.org/).