---
title: "Introduction to Shiny"
author: "Lingge Li"
date: "2/21/2016"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## What is Shiny

- A framework for building web applications

- Best for interactive data visualization

- Apps for exploratory analysis

http://shiny.datascience.uci.edu/UCIDataScienceInitiative/ClimateActionShiny/

http://shiny.datascience.uci.edu/uciMetropolitanFutures/employment_centers/

## Server and UI

- ui.R has everything you see

input widgets, plots, tables...

- server.R does the work

```{r echo=TRUE, eval=FALSE}
shinyServer(function(input, output) {
  
})

shinyUI(fluidPage(
  
))
```

## Input

- Input handled by specific widgets

- Each input has an id

- Access input value with input$id

http://shiny.rstudio.com/gallery/widget-gallery.html

## Output

- Output rendered in server with output$id

- Then displayed in ui

```{r echo=TRUE, eval=FALSE}
library(shiny)
shinyServer(function(input, output) {
  output$histogram <- renderPlot({
    hist(faithful$eruptions)
  })
})

shinyUI(fluidPage(
  plotOutput(outputId='histogram')
))
```

## Several types of output

- plotOutput (imageOutput)
- tableOutput (dataTableOutput)
- textOutput (html)
- verbatimTextOutput (console)
- htmlOutput (uiOutput)

## Datatable

- Javascript table

\includegraphics[width=300pt]{table.png}

http://shiny.rstudio.com/gallery/datatables-options.html

## MathJax

- Javascript for displaying LaTex

- Pass xtable output as raw html

```{r echo=TRUE, eval=FALSE}
output$table <- renderUI({
    M <- print(xtable(M, align=rep("c", ncol(M)+1)), 
               floating=FALSE, tabular.environment="array", 
               comment=FALSE, print.results=FALSE)
    html <- paste0("$$", M, "$$")
    list(withMathJax(), HTML(html))
})

withMathJax(),
uiOutput('table')
```

http://shiny.rstudio.com/gallery/mathjax.html

## Reactive environment

- Triggered when input changes

- Output changes accordingly

- Rendering functions reactive

## Example

```{r echo=TRUE, eval=FALSE}
library(shiny)
library(ggplot2)

shinyServer(function(input, output) {
  # read dataset
  imdb <- read.csv("~/Downloads/movie_metadata.csv")
  # subset nicolas cage movies
  cage <- imdb[imdb$actor_1_name == 'Nicolas Cage', ]
  cage$actor_1_name <- as.character(cage$actor_1_name)
  # create plot
  output$movies <- renderPlot({
    ggplot(data=cage, aes(x=title_year, y=imdb_score, label=movie_title)) +
      geom_point(alpha=0.5) + 
      geom_text(fontface='italic', size=6, vjust=1, nudge_y=0.1) + 
      labs(x='Year', y='IMDB Score') +
      geom_smooth()
  })
})
```

## Example

```{r echo=TRUE, eval=FALSE}
shinyUI(fluidPage(
  plotOutput(outputId='movies')
))
```

## Add widget

```{r echo=TRUE, eval=FALSE}
shinyUI(fluidPage(
  plotOutput(outputId='movies'),
  selectInput('color', label='Color', 
              choices=list('black', 'red', 'blue'), 
              selected='black')
))
```

## Change color

- Set colour to input$color

```{r echo=TRUE, eval=FALSE}
output$movies <- renderPlot({
  ggplot(data=cage, aes(x=title_year, y=imdb_score, label=movie_title)) +
    geom_point(alpha=0.5, colour=input$color) + 
    geom_text(fontface='italic', size=6, vjust=1, nudge_y=0.1, colour=input$color) + 
    labs(x='Year', y='IMDB Score') +
    geom_smooth()
})
```

## Slider

```{r echo=TRUE, eval=FALSE}
shinyUI(fluidPage(
  plotOutput(outputId='movies'),
  selectInput('color', label='Color', 
              choices=list('black', 'red', 'blue'), 
              selected='black'),
  sliderInput('year', label="Year", 
              min=1980, max=2016, value=c(1980, 2016))
))
```

## Filter by input$year

```{r echo=TRUE, eval=FALSE}
output$movies <- renderPlot({
  temp <- cage[cage$title_year >= input$year[1] & cage$title_year <= input$year[2], ]
  ggplot(data=temp, aes(x=title_year, y=imdb_score, label=movie_title)) +
    geom_point(alpha=0.5, colour=input$color) + 
    geom_text(fontface='italic', size=6, vjust=1, nudge_y=0.1, colour=input$color) + 
    labs(x='Year', y='IMDB Score') +
    geom_smooth()
})
```

## Layout

- Fluid grid layout (similar to bootstrap)

- 12 columns every row

- Tabset

\includegraphics[width=300pt]{grid.png}

## Rows and columns

```{r echo=TRUE, eval=FALSE}
shinyUI(fluidPage(
  fluidRow(align='center',
    plotOutput(outputId='movies')
  ),
  fluidRow(
    column(3, offset=1,
           selectInput('color', label='Color', 
                       choices=list('black', 'red', 'blue'), 
                       selected='black')),
    column(4,
           sliderInput('year', label="Year", 
                       min=1980, max=2016, value=c(1980, 2016)))
    )
))
```

## Html

- Customize html and css style files

- Shiny themes

\includegraphics[width=300pt]{theme.png}

https://rstudio.github.io/shinythemes/

## Interactive documents

- Embed interactive plots in markdown

```{r echo=TRUE, eval=FALSE}
---
title: "Nicolas Cage Movies"
output: html_document
runtime: shiny
---
```
  
http://rmarkdown.rstudio.com/authoring_shiny.html

## Shiny server

- Deploy apps to the interwebs

- http://shiny.datascience.uci.edu/server/

## Other packages

- Widgets for Javascript data visualization

http://www.htmlwidgets.org/

## Resources

- Tutorial

http://shiny.rstudio.com/tutorial/

- Cheatsheet

http://shiny.rstudio.com/images/shiny-cheatsheet.pdf

- Gallery with source code

http://shiny.rstudio.com/gallery/

