---
title: "Introduction to ggplot"
author: "Lingge Li"
date: "2/21/2017"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Why ggplot

- Beautiful aesthetics
- Flexible and powerful

```{r echo=FALSE, eval=TRUE, warning=FALSE, fig.width=4, fig.height=2.5, fig.align='center'}
library(ggplot2)
ggplot(data=iris, aes(x=Species, y=Petal.Width)) +
  geom_violin(aes(fill=Species)) +
  labs(title='Iris Petal Width')
```


## However...

- Syntax slightly compilcated at first glance

```{r echo=TRUE, eval=FALSE}
ggplot(data=iris, aes(x=Species, y=Petal.Width)) +
  geom_violin(aes(fill=Species)) +
  labs(title='Iris Petal Width')
```

## Layered grammar of graphics

- ggplot2 follows a specific grammar of graphics

\includegraphics[width=300pt]{layer.png}

*Example taken from Hadley Wickham's book*

http://vita.had.co.nz/papers/layered-grammar.pdf

## How to make a plot

- Geometric objects (geom)
- Aesthetic mapping (aes)
- Statistical transformation (stat)
- Scales and coordinate system

## Geoms

- Wide range of geometric objects from points to complex shapes

- geom_point, geom_line, geom_histogram, geom_boxplot...

- Multiple geometric objects on the same plot with +


## Aesthetics

- Coordinate positions (always needed)

- Colour, fill, shape, size...

## Data + mapping

- aes() maps a dataframe to geom

- Each geom can have its own mapping

```{r echo=TRUE, eval=FALSE}
geom_point(data, aes(x, y))
```

## Scatterplot

```{r echo=TRUE, eval=TRUE, fig.width=3, fig.height=2, fig.align='center'}
ggplot() + geom_point(data=iris, 
             aes(x=Petal.Length, y=Petal.Width))
```

- The code below equivalent and more commonly seen

```{r echo=TRUE, eval=FALSE, fig.width=4, fig.height=3, fig.align='center'}
ggplot(data=iris, aes(x=Petal.Length, y=Petal.Width)) +
  geom_point()
```

## Colour and shape

```{r echo=FALSE, eval=TRUE, fig.width=4, fig.height=2.5, fig.align='center'}
ggplot() + 
  geom_point(data=iris, aes(x=Petal.Length, y=Petal.Width, 
                          colour=Species))
```

How would you change the shape?

## Mapping vs setting

- What is the key difference

```{r echo=TRUE, eval=TRUE, fig.width=3, fig.height=2, fig.align='center'}
ggplot(data=iris, aes(x=Petal.Length, y=Petal.Width)) +
  geom_point(colour='red')
```

## Stat

- Draw a histogram the hard way

```{r echo=TRUE, eval=TRUE, fig.width=3, fig.height=2, fig.align='center'}
x <- rnorm(500)
temp <- as.data.frame(x)
ggplot(temp, aes(x=x)) + 
  geom_histogram(stat='bin', binwidth=0.1)
```

- Most plots do not require stat

## Pie chart

- Pie charts are surprisingly tricky to make

- Stacked bar chart in polar coordinates

```{r echo=TRUE, eval=TRUE, fig.width=3, fig.height=2, fig.align='center'}
ggplot(mtcars, aes(x=factor(1), fill=factor(cyl))) +
  geom_bar(width=1) + coord_polar(theta='y')
```

http://docs.ggplot2.org/current/coord_polar.html

## Quick plot

- qplot similar to base plot

- http://docs.ggplot2.org/current/qplot.html

```{r echo=TRUE, eval=TRUE, warning=FALSE, out.width='.45\\linewidth', fig.show='hold', fig.align='center'}
x <- rnorm(500)
qplot(x, geom='histogram', binwidth=0.1)
qplot(x, geom='dotplot', binwidth=0.1)
```

## Fancy plots

- Basic plots done fancily with ggplot

```{r echo=TRUE, eval=TRUE, warning=FALSE, out.width='.45\\linewidth', fig.show='hold', fig.align='center'}
ggplot(data=iris, aes(x=Species, y=Petal.Width)) +
  geom_violin(aes(fill=Species))

ggplot(data=iris, aes(x=Petal.Width)) +
  geom_density(aes(fill=Species))
```

## Violin plot with points

- geom_violin + geom_point

```{r echo=TRUE, eval=TRUE, fig.width=4, fig.height=2.5, fig.align='center'}
ggplot(data=iris, aes(x=Species, y=Petal.Width)) +
  geom_violin(aes(fill=Species)) + geom_point()
```

## Jitter

```{r echo=TRUE, eval=TRUE, fig.width=4, fig.height=2.5, fig.align='center'}
ggplot(data=iris, aes(x=Species, y=Petal.Width)) +
  geom_violin(aes(fill=Species)) + 
  geom_jitter(alpha=0.4)
```

## IMDB data

- Dataset contains over 5000 movies and 28 variables from IMDB

https://www.kaggle.com/deepmatrix/imdb-5000-movie-dataset

```{r echo=TRUE, eval=TRUE}
imdb <- read.csv("~/Downloads/movie_metadata.csv")
colnames(imdb)[1:20]
```

## IMDB score vs log10(budget)

```{r echo=TRUE, eval=TRUE}
imdb <- imdb[!is.na(imdb$imdb_score) & !is.na(imdb$budget), ]
```

```{r echo=FALSE, eval=TRUE, fig.width=4, fig.height=2.5, fig.align='center'}
ggplot(data=imdb, aes(x=log10(budget), y=imdb_score)) +
  geom_point(alpha=0.5, size=0.5)
```

## Add regression line

```{r echo=TRUE, eval=TRUE}
imdb$regression <- lm(imdb_score ~ log10(budget), data=imdb)$fitted.values
```

```{r echo=FALSE, eval=TRUE, fig.width=4, fig.height=2.5, fig.align='center'}
ggplot(data=imdb, aes(x=log10(budget), y=imdb_score)) +
  geom_point(alpha=0.5, size=0.5) + geom_line(aes(y=regression), colour='red')
```

## geom_smooth

- Kernel smoother

```{r echo=TRUE, eval=TRUE, fig.width=4, fig.height=2.5, fig.align='center'}
ggplot(data=imdb, aes(x=log10(budget), y=imdb_score)) +
  geom_point(alpha=0.5, size=0.5) + geom_smooth()
```

## Nicolas Cage Movies

```{r echo=TRUE, eval=TRUE, fig.width=4, fig.height=2.5, fig.align='center'}
cage <- imdb[imdb$actor_1_name == 'Nicolas Cage', ]
cage$actor_1_name <- as.character(cage$actor_1_name)
ggplot(data=cage, aes(x=title_year, y=imdb_score, label=movie_title)) +
  geom_point(alpha=0.5) + 
  geom_text(fontface='italic', size=2, vjust=1, nudge_y=0.1) + 
  geom_smooth()
```

## Compare with Leonardo DiCaprio

```{r echo=FALSE, eval=TRUE, fig.width=5, fig.height=3}
dicaprio <- imdb[imdb$actor_1_name == 'Leonardo DiCaprio', ]
dicaprio$actor_1_name <- as.character(dicaprio$actor_1_name)
temp <- rbind(cage, dicaprio)
ggplot(data=temp, aes(x=title_year, y=imdb_score, label=movie_title, colour=actor_1_name)) +
  geom_point() + geom_text(fontface='italic', size=2, vjust=1, nudge_y=0.1) +
  geom_smooth()
```

## Scale

- Transform the scale instead of data

- How is this different from before

```{r echo=TRUE, eval=TRUE, fig.width=4, fig.height=2.5, fig.align='center'}
ggplot(data=imdb, aes(x=budget, y=imdb_score)) +
  scale_x_log10() +
  geom_point(alpha=0.5, size=0.5) + geom_smooth()
```

## Custom tick marks and labels

```{r echo=TRUE, eval=TRUE, fig.width=4, fig.height=2.5, fig.align='center'}
ggplot(data=imdb, aes(x=budget, y=imdb_score)) +
  geom_point(alpha=0.5, size=0.5) + geom_smooth() +
  scale_x_continuous(breaks=c(1e6, 1e8), 
                     labels=c('Low budget', 'Blockbuster'),
                     trans='log10')
```

## Colour scale

```{r echo=TRUE, eval=TRUE, fig.width=4, fig.height=2, fig.align='center'}
imdb <- imdb[!is.na(imdb$gross), ]
ggplot(data=imdb, aes(x=budget, y=imdb_score)) +
  geom_point(aes(colour=gross), alpha=0.5, size=0.5) + 
  scale_x_log10() +
  scale_color_continuous(name='Box office gross', breaks = c(2e+8, 4e+8, 6e+8),
                         labels = c('2 million', '4 million', '6 million'),
                         low = 'blue', high = 'red')
```

## Putting it together

```{r echo=FALSE, eval=TRUE}
ggplot(data=imdb, aes(x=budget, y=imdb_score)) +
  geom_point(aes(colour=gross), alpha=0.5) +
  scale_x_continuous(breaks=c(1e6, 1e8), 
                     labels=c('Low', 'Blockbuster'),
                     trans='log10') +
  geom_smooth() +
  scale_color_continuous(name='Box office gross', breaks = c(2e+8, 4e+8, 6e+8),
                         labels = c('2 million', '4 million', '6 million'),
                         low = 'blue', high = 'red') +
  annotate('point', x=2e+08, y=8.3, colour='#D55E00', shape=16, size=4) +
  annotate('text', x=2e+08, y=8.7, label='Toy Story 3', fontface='italic', size=6, angle=30) +
  annotate('point', x=210000000, y=5.7, colour='#009E73', shape=17, size=4) +
  annotate('text', x=210000000, y=5.3, label='Transformers 3', fontface='italic', size=6) +
  theme_bw() +
  labs(title='IMDB Movies', x='Budget', y='Rating') + 
  theme(plot.title=element_text(size=rel(2), colour='blue')) +
  theme(legend.position=c(0.2, 0.2))
```

## Details

- Theme 

- http://docs.ggplot2.org/current/theme.html

```{r echo=TRUE, eval=FALSE}
ggplot(data=imdb, aes(x=budget, y=imdb_score)) +
  geom_point(aes(colour=gross), alpha=0.5) +
  scale_x_continuous(breaks=c(1e6, 1e8), 
                     labels=c('Low', 'Blockbuster'),
                     trans='log10') +
  geom_smooth() +
  scale_color_continuous(name='Box office gross', breaks = c(2e+8, 4e+8, 6e+8),
                         labels = c('2 million', '4 million', '6 million'),
                         low = 'blue', high = 'red') +
  annotate('point', x=2e+08, y=8.3, colour='#D55E00', shape=16, size=4) +
  annotate('text', x=2e+08, y=8.7, label='Toy Story 3', fontface='italic', size=6, angle=30) +
  annotate('point', x=210000000, y=5.7, colour='#009E73', shape=17, size=4) +
  annotate('text', x=210000000, y=5.3, label='Transformers 3', fontface='italic', size=6) +
  theme_bw() +
  labs(title='IMDB Movies', x='Budget', y='Rating') + 
  theme(plot.title=element_text(size=rel(2), colour='blue')) +
  theme(legend.position=c(0.2, 0.2))
```

## Facet

- Similar plots for different periods with facet

```{r echo=TRUE, eval=TRUE, fig.width=4, fig.height=2}
imdb$period <- cut(imdb$title_year, breaks=quantile(imdb$title_year), 
                   labels=c('Before 1999', '1999-2005', '2005-2010', '2010-2016'),
                   include.lowest=TRUE)
```

```{r echo=TRUE, eval=FALSE}
ggplot(data=imdb, aes(x=log10(budget), y=imdb_score)) + 
  geom_point(aes(colour=gross), alpha=0.5, size=0.5) +
  geom_smooth(method='lm') +
  scale_color_continuous(name='Box office gross', breaks = c(2e+8, 4e+8, 6e+8),
                         labels = c('2 million', '4 million', '6 million'),
                         low = 'blue', high = 'red') +
  facet_wrap(~period)
```

## Facet

```{r echo=FALSE, eval=TRUE}
ggplot(data=imdb, aes(x=log10(budget), y=imdb_score)) + 
  geom_point(aes(colour=gross), alpha=0.5, size=0.5) +
  geom_smooth(method='lm') +
  scale_color_continuous(name='Box office gross', breaks = c(2e+8, 4e+8, 6e+8),
                         labels = c('2 million', '4 million', '6 million'),
                         low = 'blue', high = 'red') +
  facet_wrap(~period)
```

## Multiple plots

- Multiplot function 

http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/  

- gridExtra package 

https://cran.r-project.org/web/packages/gridExtra/vignettes/arrangeGrob.html

## Resources

- Documentation 

http://docs.ggplot2.org/current/

- Cheatsheet 

https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf

- R Cookbook 

http://www.cookbook-r.com/Graphs/

- Stackoverflow 

http://stackoverflow.com/questions/tagged/ggplot2