---
title: "Data Visualization with ggplot"
author: "Lingge Li"
date: "10/19/2018"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Why ggplot

- Beautiful aesthetics
- Flexible and powerful

```{r echo=FALSE, eval=TRUE, warning=FALSE, fig.width=4, fig.height=2.5, fig.align='center'}
library(data.table)
library(ggplot2)
library(gridExtra)
library(ggthemes)

url <-'https://raw.githubusercontent.com/collnell/burritos/master/sd_burritos.csv'
burritos <- fread(url)

burritos$Guac[is.na(burritos$Guac)] <- 0
burritos$Guac <- ifelse(burritos$Guac == 1, 'Guac', 'No Guac')

ggplot(burritos) + 
  geom_point(aes(x=Salsa, y=Taste, color=Guac)) + 
  geom_smooth(aes(x=Salsa, y=Taste, color=Guac), 
              method="loess") +
  scale_colour_manual(values=c("darkgreen", "red")) +
  ggtitle('What Makes A Burrito Taste Good') +
  theme_economist()
```


## However...

- Syntax slightly compilcated at first glance

```{r echo=TRUE, eval=FALSE}
ggplot(burritos) + 
  geom_point(aes(x=Salsa, y=Taste, color=Guac)) + 
  geom_smooth(aes(x=Salsa, y=Taste, color=Guac), 
              method="loess") +
  scale_colour_manual(values=c("darkgreen", "red")) +
  ggtitle('What Makes A Burrito Taste Good') +
  theme_economist()
```

## Layered grammar of graphics

- ggplot2 follows a specific grammar of graphics

\includegraphics[width=300pt]{layer.png}

*Example taken from Hadley Wickham's book*

http://vita.had.co.nz/papers/layered-grammar.pdf

## How to make a plot

- Geometric objects (geom)
- Aesthetic mapping (aes)
- Statistical transformation (stat)
- Scales and coordinate system

## Geoms

- Wide range of geometric objects from points to complex shapes

- geom_point, geom_line, geom_histogram, geom_boxplot...

- Multiple geometric objects on the same plot with +

- https://www.rstudio.com/wp-content/uploads/2016/11/ggplot2-cheatsheet-2.1.pdf


## Aesthetics

- Coordinate positions (always needed)

- Colour, fill, shape, size...

## Data + mapping

- aes() maps a dataframe to geom

- Each geom can have its own mapping

```{r echo=TRUE, eval=FALSE}
geom_point(data, aes(x, y))
```


## Stat

- Plotting distributions needs statistical transformation (count for histogram)

- Variables are often transformed to be meaningful (log for concentration)

- Statistical models can highlight data patterns (regression line)


## Scales and coordinate system

- Axis ticks and labels can be customized

- Color scale can also be modified

- Polar coordinates are used for pie charts


## Burritos 

- Dataset of burritos in San Diego

\includegraphics[width=300pt]{burritos.png}

https://srcole.github.io/100burritos/


## Read data

```{r echo=TRUE, eval=FALSE}
library(data.table)
library(ggplot2)

url <- "https://raw.githubusercontent.com/collnell/
burritos/master/sd_burritos.csv"
burritos <- fread(url)
```


## Univariate plot for continuous variable

```{r echo=TRUE, eval=TRUE, warning=FALSE, fig.width=4, fig.height=2.5, fig.align='center'}
ggplot(data=burritos) + 
  geom_histogram(aes(x=Cost), bins=20)
```

geom_density


## Univariate plot for discrete variable

```{r echo=TRUE, eval=TRUE, warning=FALSE, fig.width=4, fig.height=2.5, fig.align='center'}
burritos$Burrito <- factor(burritos$Burrito)
#summary(burritos$Burrito)
top.burritos <- c('al pastor', 'california', 'carne asada', 'carnitas', 'surf & turf')
temp <- burritos[burritos$Burrito %in% top.burritos, ]

ggplot() + geom_bar(data=temp, aes(x=Burrito))
```


## Pie chart

- Pie charts are surprisingly tricky to make

- Stacked bar chart in polar coordinates

```{r echo=TRUE, eval=TRUE, warning=FALSE, fig.width=4, fig.height=2.5, fig.align='center'}
ggplot(temp, aes(x=factor(1), fill=Burrito)) +
  geom_bar(width=1) + coord_polar(theta='y')
```


## Quick plot

- qplot similar to base plot

- http://docs.ggplot2.org/current/qplot.html

```{r echo=TRUE, eval=TRUE, warning=FALSE, out.width='.45\\linewidth', fig.show='hold', fig.align='center'}
x <- rnorm(500)
qplot(burritos$Cost, geom='histogram')
```


## Bivariate plot

```{r echo=TRUE, eval=TRUE, warning=FALSE, fig.width=4, fig.height=2.5, fig.align='center'}
ggplot() + geom_violin(data=burritos, aes(x=Chips, y=Cost))
#ggplot() + geom_boxplot(data=burritos, aes(x=Chips, y=Cost))
```


## Mapping vs setting

- What is the difference

```{r echo=TRUE, eval=FALSE}
ggplot() + 
  geom_violin(data=burritos, aes(x=Chips, y=Cost), fill='red')
ggplot() + 
  geom_violin(data=burritos, aes(x=Chips, y=Cost, fill=Chips))
```

- Try this with histogram


## Violin plot with points

- geom_violin + geom_point

- geom_jitter

```{r echo=TRUE, eval=TRUE, fig.width=4, fig.height=2.5, fig.align='center'}
ggplot(data=burritos) + geom_violin(aes(x=Chips, y=Cost, fill=Chips)) +
  geom_point(aes(x=Chips, y=Cost))
```


## Changing the scale

```{r echo=TRUE, eval=TRUE, fig.width=4, fig.height=2.5, fig.align='center'}

#ggplot(burritos) + geom_point(aes(x=Volume, y=Cost))
#ggplot(burritos) + geom_point(aes(x=log10(Volume), y=Cost))
ggplot(burritos) + geom_point(aes(x=Volume, y=Cost)) +
  scale_x_log10()
```

## Adding trend

- Linear regression or kernel smoother

```{r echo=TRUE, eval=TRUE, fig.width=4, fig.height=2.5, fig.align='center'}
ggplot(burritos) + geom_point(aes(x=Volume, y=Cost)) +
  geom_smooth(method="lm", aes(x=Volume, y=Cost)) +
  scale_x_log10()
```


## Visualize three variables

- Use color or shape to indicate a third variable

```{r echo=TRUE, eval=TRUE, warning=FALSE, fig.width=4, fig.height=2.5, fig.align='center'}
#burritos$Guac[is.na(burritos$Guac)] <- 0
#burritos$Guac <- ifelse(burritos$Guac == 1, 'Guac', 'No Guac')

ggplot(burritos) + 
  geom_point(aes(x=Salsa, y=Taste, color=Guac)) + 
  geom_smooth(aes(x=Salsa, y=Taste, color=Guac))

#ggplot(burritos) + 
#  geom_point(aes(x=Salsa, y=Taste, color=Cost))
```


## More than three variables

- Similar plots for different categories with facet

- http://www.cookbook-r.com/Graphs/Facets_(ggplot2)/

```{r echo=TRUE, eval=TRUE, warning=FALSE, fig.width=4, fig.height=2.5, fig.align='center'}
ggplot(temp) + 
  geom_point(aes(x=Salsa, y=Taste, color=Guac)) + 
  facet_wrap(~Burrito)
```


## Multiple plots

```{r echo=FALSE, eval=TRUE, warning=FALSE, fig.width=4, fig.height=2.5, fig.align='center'}
p1 <- ggplot(burritos) + 
  geom_point(aes(x=Salsa, y=Taste)) +
  geom_smooth(aes(x=Salsa, y=Taste), method="loess")
p2 <- ggplot(burritos) + 
  geom_point(aes(x=Hunger, y=Taste)) +
  geom_smooth(aes(x=Hunger, y=Taste), method="loess")
p3 <- ggplot(burritos) + 
  geom_point(aes(x=Tortilla, y=Taste)) +
  geom_smooth(aes(x=Tortilla, y=Taste), method="loess")
p4 <- ggplot(burritos) + 
  geom_point(aes(x=Meat, y=Taste)) +
  geom_smooth(aes(x=Meat, y=Taste), method="loess")
grid.arrange(p1, p2, p3, p4, ncol=2)
```


## gridExtra

- https://cran.r-project.org/web/packages/gridExtra/vignettes/arrangeGrob.html

- Each plot is an object

```{r echo=TRUE, eval=FALSE, warning=FALSE, fig.width=4, fig.height=2.5, fig.align='center'}
library(gridExtra)

p1 <- ggplot(burritos) + geom_point(aes(x=Salsa, y=Taste)) +
  geom_smooth(aes(x=Salsa, y=Taste), method="loess")
p2 <- ggplot(burritos) + geom_point(aes(x=Hunger, y=Taste)) +
  geom_smooth(aes(x=Hunger, y=Taste), method="loess")
p3 <- ggplot(burritos) + geom_point(aes(x=Tortilla, y=Taste)) +
  geom_smooth(aes(x=Tortilla, y=Taste), method="loess")
p4 <- ggplot(burritos) + geom_point(aes(x=Meat, y=Taste)) +
  geom_smooth(aes(x=Meat, y=Taste), method="loess")

grid.arrange(p1, p2, p3, p4, ncol=2)
```


## Custom tick marks and labels

```{r echo=TRUE, eval=TRUE, warning=FALSE, fig.width=4, fig.height=2.5, fig.align='center'}
ggplot(burritos) + 
  geom_point(aes(x=Salsa, y=Taste)) +
  scale_y_continuous(breaks=c(1, 2, 3, 4, 5),
                     labels=c('Throw away', 'Edible', 
                              'Okay', 'Decent', 'Awesome')) +
  ylab('Deliciousness')
```

## Color scale

- scale_color_continuous

- scale_colour_manual

```{r echo=FALSE, eval=TRUE, warning=FALSE, fig.width=4, fig.height=2.5, fig.align='center'}
ggplot(burritos) + 
  geom_point(aes(x=Salsa, y=Taste, color=Cost)) +
  scale_color_continuous(name='Price', breaks = c(4, 6, 8, 10),
                       labels = c('$4', '$6', '$8', '$10'),
                       low = 'blue', high = 'red')
```


## Other details

- annotate

- theme

```{r echo=TRUE, eval=FALSE, warning=FALSE, fig.width=4, fig.height=2.5, fig.align='center'}
ggplot(burritos) + 
  geom_point(aes(x=Salsa, y=Taste, color=Cost)) + 
  annotate('text', x=3, y=5.3, label='Best in San Diego', 
           fontface='italic', size=6, colour='blue') +
  theme_bw() +
  labs(title='Burritos', x='Tomato-based Mexican Sauce', y='Deliciousness') + 
  theme(plot.title=element_text(size=rel(2), hjust = 0.5)) +
  theme(legend.position=c(0.1, 0.3))
```


## Other details

```{r echo=FALSE, eval=TRUE, warning=FALSE, fig.width=4, fig.height=2.5, fig.align='center'}
ggplot(burritos) + 
  geom_point(aes(x=Salsa, y=Taste, color=Cost)) + 
  annotate('text', x=3, y=5.3, label='Best in San Diego', 
           fontface='italic', size=6, colour='blue') +
  theme_bw() +
  labs(title='Burritos', x='Tomato-based Mexican Sauce', y='Deliciousness') + 
  theme(plot.title=element_text(size=rel(2), hjust = 0.5)) +
  theme(legend.position=c(0.1, 0.3))
```


## ggthemes and other extensions

- The Economist and FiveThirtyEight themes

- http://www.ggplot2-exts.org/ggthemes.html


## Resources

- Documentation 

http://docs.ggplot2.org/current/

- R Cookbook 

http://www.cookbook-r.com/Graphs/

- Stackoverflow 

http://stackoverflow.com/questions/tagged/ggplot2


## Polygon

```{r echo=TRUE, eval=TRUE, fig.width=4, fig.height=2, fig.align='center'}
x <- c(1, 2, 3, 4, 2.5)
y <- c(3, 0, 1, 4, 4)
variable <- c(1, 2, 1, 2, 2)
example <- data.frame(x, y, variable)

ggplot(data=example) +
  geom_point(aes(x=x, y=y, colour=as.factor(variable))) +
  geom_polygon(aes(x=x, y=y), colour='black', fill=NA)
```


## Polygons

```{r echo=TRUE, eval=TRUE}
x <- c(5, 6, 4)
y <- c(4, 3, 1)
variable <- c(3, 3, 3)
triangle <- data.frame(x, y, variable)
```

```{r echo=TRUE, eval=TRUE, fig.width=4, fig.height=2, fig.align='center'}
example$group <- 1
triangle$group <- 2
both <- rbind(example, triangle)
ggplot(data=both) +
  geom_point(aes(x=x, y=y, colour=as.factor(variable))) +
  geom_polygon(aes(x=x, y=y, group=group), colour='black', fill=NA)
```


## State dataframe

```{r echo=TRUE, eval=TRUE, message=FALSE}
library(maps)

states <- map_data('state')
head(states)
```


## California

```{r echo=TRUE, eval=TRUE, message=FALSE, fig.width=3, fig.height=3, fig.align='center'}
california <- states[states$region == 'california', ]
ggplot(data=california) + 
  geom_polygon(aes(x=long, y=lat), fill='orange', color='red')
```


## USA

```{r echo=TRUE, eval=TRUE, message=FALSE, fig.width=4, fig.height=3, fig.align='center'}
ggplot(data=states) + 
  geom_polygon(aes(x=long, y=lat, group=group, fill=region), show.legend=FALSE)
```


## Shapefile

- Packages in R can read and process common shapefiles

```{r echo=TRUE, eval=FALSE}
CAPD <- readOGR('.', 'CPAD_2016b1_SuperUnits')
CAPD <- spTransform(CAPD, 
CRS("+proj=longlat +ellps=WGS84 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs"))
proj4string(CAPD) <- CRS("+proj=longlat +ellps=WGS84 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs")
```


## ggmap and Los Angeles Metro Bike Share Trip Data

- https://github.com/dkahle/ggmap

- https://www.kaggle.com/cityofLA/los-angeles-metro-bike-share-trip-data/version/25#metro-bike-share-trip-data.csv

```{r echo=FALSE, eval=TRUE, message=FALSE, fig.width=4, fig.height=3, fig.align='center'}
library(ggmap)

metro.bike.share.trip.data <- read.csv("~/Downloads/metro-bike-share-trip-data.csv")
metro.bike.share.trip.data <- metro.bike.share.trip.data[metro.bike.share.trip.data$Starting.Station.Latitude > 0, ]

# use qmplot to make a scatterplot on a map
qmplot(Starting.Station.Longitude, Starting.Station.Latitude, 
       data=metro.bike.share.trip.data[1:500, ], maptype="toner-lite", color=I("red"))
```


## Plot trips as lines

```{r echo=TRUE, eval=TRUE, message=FALSE, fig.width=4, fig.height=3, fig.align='center'}
qmplot(Starting.Station.Longitude, Starting.Station.Latitude, 
       data=metro.bike.share.trip.data[1:500, ], maptype="toner-lite", color=I("red")) +
  geom_segment(aes(x=Starting.Station.Longitude, xend=Ending.Station.Longitude,
                   y=Starting.Station.Latitude, yend=Ending.Station.Latitude),
               color='red', alpha=0.2)
```


## The end

- https://rstudio.github.io/leaflet/

- https://shiny.rstudio.com/gallery/

- https://plot.ly/r/

- https://rstudio.github.io/r2d3/articles/introduction.html

- Thank you!