---
title       : R-Markdown
subtitle    : UCI Data Science Initiative
author      : Sepehr Akhavan
job         : Dept. of Statistics
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : mathjax            # {mathjax, quiz, bootstrap}
logo     : logo.png
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
github:
  user: UCIDataScienceInitiative
  repo: AdvancedRWorkshop
---

## Introduction

```{r,echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
library(ggplot2)
library(rmarkdown)
library(knitr)
```

+ Markdown is a simple formatting language designed to make authoring content easy for everyone.

+ Rather than writing complex markup code (e.g. HTML or LaTeX), Markdown enables the use of a syntax much more like plain-text email.

+ R Markdown is an authoring format that enables easy creation of dynamic documents (reports, presentations, ...)

+ R Markdown combines markdown (an easy-to-write plain text format) with R

+ R Markdown documents are fully reproducible 

---

##  Installation

+ We need to install two packages: knitr, rmarkdown

+ To install these packages, you can simply run:
```{r, eval=FALSE}
install.packages("rmarkdown")
install.packages("knitr")
```

+ Once installed, run:
```{r, eval=FALSE}
library(knitr)
library(rmarkdown)
```

+ And let's open a markdown document

---

## Where to learn?

+ Check out the R-Markdown Cheat Sheet:
  + http://shiny.rstudio.com/articles/rm-cheatsheet.html
  
+ Check out knitr website:
  + http://yihui.name/knitr/

+ R Markdown Reference Guide:
  + https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf


---

## R-Markdown - Output formats:

+ By using [pandoc] (http://pandoc.org), you can have different outputs:

  + html_document
  + pdf_document
  + word_document
  + beamer_document
  + ioslides_presentation


---

## R-Markdown - How to Embed R codes?

+ inline code: Surround code with back ticks and r.

+ code chunks: 

+ R chunks may include some options:
  + eval
  + echo
  + warning, error, message
  + tidy: to reformat code in a tidy way (default is FALSE)
  + results: "markup", "asis", "hold", "hide"
  + comment: Comment character to preface results with ("##")
  + fig.width, fig.height (in inches)


---

## Interactive Documents:

+ With R Markdown, you can even report an interactive shiny document.

+ To do so:
  + 1) keep output as html_output (or ioslides_presentation)
  + 2) add shiny run time to YAML: runtime: shiny
  + 3) Need Shiny input functions and Shiny render functions

+ More info: http://rmarkdown.rstudio.com/authoring_shiny.html


---

## Embeding Equations:

  + One "dollar" sign for inline equation
  
  + Two "dollar" signs for display equation


---

## Let's have an example in RMarkdown:
