---
title: "Final Problems"
author: "Lingge Li"
output: pdf_document
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(alr4)
```

## Week 7

Heights of 16 Gothic cathedrals are listed below.

```{r echo=FALSE, eval=TRUE, warning=FALSE}
cathedral[cathedral$Type=='Gothic', 1:2]
```

1. Calculate the sample mean and standard deviation. Then sketch the sampling distribution of the mean.

2. Construct a 95\% confidence interval of the mean.

3. Interpret the confidence interval in the context of the problem.

Source: S. Weisberg 2014 Applied Linear Regression

## Week 8

Two experiments were performed to test if manipulating the air conditioning fans in the Minneapolis metrodome can effect the distance travelled by a baseball in 2003. The summary of travel distances is given below. 

**Headwind** mean 356.7 standard deviation 8.4 $n=17$

**Tailwind** mean 350.8 standard deviation 10.3 $n=17$

1. Write the null hypothesis and alternative hypothesis.

2. State the appropriate test and perform it.

2. Do baseballs travel farther in tailwind condition than in headwind condition? Justify.

Source: http://finzi.psych.upenn.edu/library/alr3/html/domedata.html

## Week 9

Ratings of 364 professors (from 1 to 5) at one large campus in the Midwest were collected in 2011 from Humanities, Social Sciences, STEM and Professional disciplines. We are interested to know whether the professor ratings differ across disciplines.

1. Write the null hypothesis and alternative hypothesis.

2. Based on the ANOVA table, state the test result.

```{r echo=FALSE, eval=TRUE}
fit <- lm(quality~discipline, data=Rateprof)
anova(fit)
```

3. Interpret the result in the context of the problem.

```{r echo=FALSE, eval=TRUE, fig.width=4, fig.height=3}
qplot(discipline, quality, data=Rateprof, geom='boxplot')
```

4. Is the result consistent with the side-by-side boxplot? Explain.

Source: http://pareonline.net/getvn.asp?v=16&n=18

