---
title: "Shiny Workshop"
author: "Lingge Li"
date: "12/2/2016"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## What is Shiny

- A web framework for building applications that run in R

- Great for interactive data visualization

- Apps for education, data visualization and analysis

## Server and UI

- ui.R defines everything you see

input widgets, plots, tables...

- server.R does the work

create plots, tables...

```{r echo=TRUE, eval=FALSE}
shinyServer(function(input, output) {
  
})

shinyUI(fluidPage(
  
))
```

## Input

- Input is handled by specific widgets

- Each input has an id

- Can be accessed with input$id

http://shiny.rstudio.com/gallery/widget-gallery.html

## Output

- Output is rendered in server as output$id

- Then displayed by id in ui

```{r echo=TRUE, eval=FALSE}
shinyServer(function(input, output) {
  output$main_plot <- renderPlot({
    hist(faithful$eruptions)
  })
})

shinyUI(fluidPage(
  plotOutput(outputId = "main_plot")
))
```

## Several types of output

- htmlOutput (uiOutput)
- plotOutput (imageOutput)
- tableOutput (dataTableOutput)
- textOutput (html)
- verbatimTextOutput (console)

## Reactive environment

- Triggered when input changes

- Change output accordingly

- Rendering functions reactive

## MoMA example

```{r echo=TRUE, eval=FALSE}
library(shiny)
library(ggplot2)

shinyServer(function(input, output) {
  Paintings <- read.csv("Paintings.csv")
  Paintings <- Paintings[Paintings$Height..cm.<500,]
  Paintings <- Paintings[Paintings$Width..cm.<1000,]
  Paintings <- Paintings[!is.na(Paintings$Height..cm.),]
  Paintings <- Paintings[!is.na(Paintings$Width..cm.),]

  output$painting <- renderPlot({
    ggplot(data=Paintings, 
           aes(x=Width..cm., y=Height..cm.)) + geom_point() +
      ggtitle('MoMA Painting Dimensions')
  })
})
```

## Select color

- How can we use input$color in server

```{r echo=TRUE, eval=FALSE}
shinyUI(fluidPage(
  mainPanel(plotOutput('painting')),
  selectInput('color', label = 'Point color', 
              choices = list('black', 'red', 'blue'), 
              selected = 'black')
))
```

## Only Picasso

- Maybe extend this with selectInput

```{r echo=TRUE, eval=FALSE}
checkboxInput('picasso', label="Only Picasso", value=FALSE)

output$painting <- renderPlot({
    if (input$picasso) {
      dat <- Paintings[Paintings$Artist=='Pablo Picasso',]
    } else {
      dat <- Paintings
    }
    ...
  })
```

## Slider

- Filter by year acquired

```{r echo=TRUE, eval=FALSE}
sliderInput('year', label=h3("Year acquired"), 
            min=1930, max=2016, value=c(1950, 2000))
```

## Mouse events

- Click, double click, hover, brush

```{r echo=TRUE, eval=FALSE}
plotOutput('painting', 
           click = "plot_click",
           dblclick = "plot_dblclick",
           hover = "plot_hover",
           brush = brushOpts(id = "plot_brush", 
                             resetOnNew = TRUE))
```

## Click info

- Print mouse events

```{r echo=TRUE, eval=FALSE}
output$click_info <- renderPrint({
    str(input$plot_click)
})

verbatimTextOutput("click_info")
```

## Double click

- Observe double click

- Reactive values

```{r echo=TRUE, eval=FALSE}
ranges <- reactiveValues(x = NULL, y = NULL)

observeEvent(input$plot_dblclick, {
    brush <- input$plot_brush
    if (!is.null(brush)) {
      ranges$x <- c(brush$xmin, brush$xmax)
      ranges$y <- c(brush$ymin, brush$ymax)
    } else {
      ranges$x <- NULL
      ranges$y <- NULL
    }
})
```

## Zoom

- Change plot coordinate system

```{r echo=TRUE, eval=FALSE}
+ coord_cartesian(xlim = ranges$x, ylim = ranges$y)
```

## Datatable

- Popular javascript table

\includegraphics[width=300pt]{table.png}

http://shiny.rstudio.com/gallery/datatables-options.html

## MathJax

- Javascript for displaying LaTex

- Pass xtable output as raw html

```{r echo=TRUE, eval=FALSE}
output$table <- renderUI({
    M <- print(xtable(M, align=rep("c", ncol(M)+1)), 
               floating=FALSE, tabular.environment="array", 
               comment=FALSE, print.results=FALSE)
    html <- paste0("$$", M, "$$")
    list(withMathJax(), HTML(html))
})

withMathJax(),
uiOutput('table')
```

http://shiny.rstudio.com/gallery/mathjax.html

## Layout

- Fluid grid layout (similar to bootstrap)

- 12 columns every row

- Tabset

\includegraphics[width=300pt]{grid.png}

## Html

- Customize html and css style files

- Shiny themes

\includegraphics[width=300pt]{theme.png}

https://rstudio.github.io/shinythemes/

## Shiny server

- Deploy apps to the interwebs

- http://shiny.datascience.uci.edu/server/

## Other packages

- Framework for Javascript visualizations

http://www.htmlwidgets.org/

## Extra

- Cheatsheet

http://shiny.rstudio.com/images/shiny-cheatsheet.pdf

- Gallery with source code

http://shiny.rstudio.com/gallery/

## Challenge

- How can you show the name of selected artwork?

- nearPoints()

http://shiny.rstudio.com/articles/selecting-rows-of-data.html

