---
title: "An Introduction to Stringr and Regular Expressions"
author: "Brian Espinoza"
output: ioslides_presentation
---
```{r echo=F}
library(stringr)
```
# Introduction

## Introduction

- String processing is an important aspect of data mining and text analysis. 
- Though R has some built-in functions to assist, they can be bulky, slow, and incomplete...

Fret not, stringr is here!


## stringr

- stringr is a lightweight package designed by Hadley Wickham to assist with string manipulation.
- Interacts seamlessly with the pipe ( %>% ) operator from dplyr / magrittr.
- Much like Hadley's other packages, stringr's function names are consistent and its arguments are easy to understand.

## Before We Begin

### **Review of Strings**

- Character strings in R are wrapped with quotes " "
  - Character strings can be letters "a", 
                             numbers "1", 
                             symbols "&", 
                             or both "a1&"
- While numbers can be both integers and characters, letters and symbols have no integer meaning and thus create NAs.
```{r}
as.integer(c("a", "&"))
```

## Before We Begin

### **Review of Strings**
- Concatenating strings and integers with the c() function will convert the integers to characters.
- By default, R converts objects to their lowest denomination.
  - Factors reduce to integers and integers reduce to characters
```{r}
c(factor("a"), "b", "&", 1)
c(as.character(factor("a")), "b", "&", 1)
```

## Agenda

- Getting Started with stringr
- Basic String Operators
- Regular Expressions
- Pattern Matching Functions
- Final Exercise

## Getting Started

#### Loading stringr

```{r warning = F, message = F, eval = F}
install.packages("stringr")
library(stringr)
```

#### Open Variables.R and load strings, fruit, and movie_titles

```{r small = T}
movie_titles <- c("gold diggers of broadway", "gone baby gone", 
    "gone in 60 seconds", "gone with the wind", "good girl, the", 
    "good burger", "goodbye girl, the", "good bye lenin!", 
    "goodfellas", "good luck chuck", "good morning, vietnam", 
    "good night, and good luck.", "good son, the", "good will hunting")

strings <- c(" 219 733 8965", "329-293-8753 ", "banana", "595 794 7569",
  "387 287 6718", "apple", "233.398.9187  ", "482 952 3315", 
  "239 923 8115 and 842 566 4692", "Work: 579-499-7527", "$1000", "Home: 543.355.3679")

fruit <- c("apple", "banana", "pear", "pineapple")

```

# Basic String Operators

## Basic String Operators

- String operators are basic string manipulation functions
- Many of them have equivalent base R functions that are much slower and bulkier

## Basic String Operators
#### **str_to_upper(string)**
    - converts strings to uppercase
      - ex. Convert all movie_titles to uppercase and store them as movie_titles

#### **str_to_lower(string)**
    - converts strings to lowercase
      - ex. Convert all movie_titles back to lowercase and save as movie_titles

#### **str_to_title(string)**
    - converts strings to title case
      - ex. Convert all movie_titles to titlecase and store them as movie_titles
      
## str_to_upper(string)
```{r}
movie_titles <- str_to_upper(movie_titles)
movie_titles
```

## str_to_lower(string)
```{r}
movie_titles <- str_to_lower(movie_titles)
movie_titles
```

## str_to_title(string)
```{r}
movie_titles <- str_to_title(movie_titles)
movie_titles
```

## Basic String Operators
#### **str_c(string, sep = "")**
    - Joins together multiple strings including integers
    - Is the stringr equivalent to paste(sep = "") or paste0()

#### **str_length(string)**
    - Returns the string length
    - Similar to base function nchar()
    - str_length() converts factors to strings and also preserves NA's
```{r small = T}
nchar(NA)
str_length(NA)
```


## Basic Strings Operators

#### **str_sub(string, start, end) <- value**
    - Subsets text within a string or vector of strings by specifying
      start and end positions.
    - Base equivalent function is substr()
    - By default, end goes to the end of the word
```{r}
fruit
str_sub(fruit, start = 3)
```

## Basic Strings Operators

#### **str_dup(string, times)**
    - Duplicates strings by a number of times.
    - Essentially copy / paste function
```{r}
str_dup(fruit, 3)
```

## Basic Strings Operators

#### **str_trim(string, side = c("both", "left", "right"))**
    - Removes leading and trailing whitespaces
    - Side argument defaults to "both"
    - ex. trim the whitespace from both sides of every string in "strings"

#### **str_pad(string, width, side = c("left", "both", "right"), pad = " ")**
    - Pads strings with whitespace to make them a certain length
    - Width argument lets users specify the width of the padding
    - Side argument defaults to "left"
    - ex. pad "movie_titles" with whitespace to the right such that each title 
      becomes 30 characters long.

## **str_trim(string, side = c("both", "left", "right"))**
#### **ex. trim the whitespace from both sides of every string in strings**
```{r small = T}
str_trim(strings)
```

## str_pad(string, width, side = c("left", "both", "right"), pad = " ")

#### **ex. pad movie_titles with whitespace to the right so that each title becomes 30 characters long.**
```{r small = T}
str_pad(movie_titles, side = "right", 30)
```


# Regular Expressions

## Regular Expressions

Pattern matching functions use patterns, otherwise known as "regular expressions" or "regex", to identify specific characteristics in strings.

#### **Common expressions:**
    - "a"  = is the letter "a"
    - "^a" = starts with the letter "a"
    - "a$" = ends with the letter "a"
    - "[ ]" = contains any letter (or number) within the brackets
    - "[ - ]" = contains any letter (or number) within this range
    - "[^ae]" = everything except these letters (or numbers)
    - "{3}" = repeat the last regex 3 times.
  
For more expressions or examples, refer to <http://www.regular-expressions.info/refquick.html>

## Compound Expressions

    Regular expressions can be combined to form compound expressions.

    - "a"  = is the letter "a"
    - "^a" = starts with the letter "a"
    - "a$" = ends with the letter "a"
    - "[ ]" = contains any letter (or number) within the brackets
    - "[ - ]" = contains any letter (or number) within this range
    - "[^ae]" = everything except these letters (or numbers)
    - "{3}" = repeat the last regex 3 times.

#### **California license plate:**
    - California plates start with a number, followed by 3 letters, followed by   
      another 3 numbers.
    - Regex expression: "^[0-9][A-Z]{3}[0-9]{3}$"

## Compound Expressions

    - "a"  = is the letter "a"
    - "^a" = starts with the letter "a"
    - "a$" = ends with the letter "a"
    - "[ ]" = contains any letter (or number) within the brackets
    - "[ - ]" = contains any letter (or number) within this range
    - "[^ae]" = everything except these letters (or numbers)
    - "{3}" = repeat the last regex 3 times.
    
  - **Your turn:** create a regex expression that would identify any social security number. Please do not write your own...
    - Format: SSS-SS-SSSS where S is any number between 0 and 9
    
## Social Security Example

  - Format: SSS-SS-SSSS where S is any number between 0 and 9
  - Regex expression: 
    - "\^[0-9]{3}-[0-9]{2}-[0-9]{4}$"



# Pattern Matching Functions

## Pattern Matching Functions

  - Now that we know how to build regular expressions, we can leverage these skills to perform even more advanced functions.
  - Pattern matching functions in stringr take advantage of the regex syntax to perform helpful tasks.
  - The usual form of these pattern matching functions consists of:
      - function(string, pattern)
    - string = a character string or a vector of character strings
    - pattern = your regex request

## Pattern Matching Functions

#### **str_detect(string, pattern)**
    - detects the presence of a pattern within a string or vector of strings
    - returns a boolean (TRUE FALSE) vector
```{r}
str_detect(fruit, pattern = "^apple$")
fruit[str_detect(fruit, "^apple$")]
```


## Pattern Matching Functions

#### **str_locate(string, pattern)**
    - locates and returns the start and end position of the first instance 
      of the pattern.
       - to locate every instance, use str_locate_all(string, pattern)
```{r}
fruit
# on the second word, this pattern exists from the first character to the sixth
str_locate(fruit, "^b[an]{5}")
```

## Pattern Matching Functions

#### **str_extract(string, pattern) or str_extract_all()**
    - matches the exact pattern to the string
    - mainly used to extract compound patterns
    
#### **str_match(string, pattern) or str_match_all()**
    - equivalent to str_extract except that str_match returns a matrix.
    - str_(m)atch(): remember "m" for matrix!

```{r small = T}
labels <- c("a99", "a92", "a93l", "b99", "b92", "b93l",
            "c99", "c92", "c93l", "e99", "e92", "e93l")

# extract everything that begins with an "a" or "e" and ends with two numbers
str_extract(labels, "^[ae][0-9]{2}$")
```


## Pattern Matching Functions

#### Exercise:
  - Extract every phone number from the variable "strings" that is composed of spaces " ". Extract it in matrix form.
```{r}
strings
```

## Pattern Matching Functions
#### Exercise:
```{r}
str_match_all(strings, pattern = "[1-9]{3} [1-9]{3} [1-9]{4}") %>% 
  unlist() %>% matrix()
```

## Pattern Matching Functions
#### Exercise:
```{r}
str_match_all(strings, pattern = "[1-9]{3} [1-9]{3} [1-9]{4}")
```

## Pattern Matching Functions
#### Exercise:
```{r}
str_match_all(strings, pattern = "[1-9]{3} [1-9]{3} [1-9]{4}") %>% 
  unlist()
str_match_all(strings, pattern = "[1-9]{3} [1-9]{3} [1-9]{4}") %>% 
  unlist() %>% matrix()
```

## Pattern Matching Functions

#### **str_replace(string, pattern, replacement)**
    - replaces the first instance of the matched pattern with the replacement string
    - str_replace_all replaces all instances of the pattern with the replacement string
    - str_replace_na replaces all NA with "NA".
```{r small = T}
str_replace(fruit, pattern = "a", replacement = "e") # only the first instance
str_replace_all(fruit, pattern = "a", replacement = "e") # every instance
```
  **Your turn:** In movie_titles, replace all instances of "Good" with "Bad".
    

## **str_replace(string, pattern, replacement)**
```{r}
str_replace_all(movie_titles, pattern = "Good", replacement = "Bad")
```

## Pattern Matching Functions
#### **str_split(string, pattern)**
  - splits a string into a variable number of pieces and returns a **list** of character vectors.
```{r}
str_split(movie_titles, "[ ,]")
```

## Pattern Matching Functions
#### **str_split_fixed(string, pattern, n)**
  - splits the string into a fixed number of pieces and returns a character **matrix**. 
```{r}
str_split_fixed(movie_titles, "[ ,]", 5)
```

# Questions?

# Final Exercise

## Final Exercise

#### **Instructions**

  1. Extract **all** phone numbers from the variable **strings**
  2. Remove all "-" and "."
  3. Split the numbers into a matrix
      - First column contains area codes and the second column contains the rest of the phone number.
  
  Matrix should contain 10 phone numbers (rows) and 2 columns
  
## Final Exercise

```{r}
strings %>% str_match_all(pattern = "[0-9]{3}[-. ][0-9]{3}[-. ][0-9]{4}") %>% 
  unlist() %>% str_replace_all(pattern = "[-. ]", replacement = " ") %>% 
  str_split_fixed(pattern = "[-. ]", 2)
```

# Thank you